One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-02-06 paper/model.tex
paper accepted, minor corrections

\section{Governing equations}\label{model} 
We consider an inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$q$ is the imposed temperature gradient since, for an ideal gas the
temperature is proportional to $c_s^2$. {\bf For convenience we will
  refer to $c_s^2$ as the disc temperature.} The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
 disc with constant $H/R$.

@@

{\bf We describe a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, in this section we consider 2D discs. 
}

@@

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
  discuss} some
general properties of the linear perturbations.

@@

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state {\bf in the perturbed disc}. {\bf  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}

@@

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and $\tbg$ is {\bf also} positive (negative),
then the local angular momentum density of the linear disturbance
{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.

{\bf 
  We demonstrate instability for low-frequency modes 
  ($|\omega|\ll m\Omega$) by explicitly showing its 
  angular momentum density 
  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
  appropriate perturbations and radial temperature gradients. 
}

\subsection{Angular momentum density of  non-axisymmetric low-frequency modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].\label{ang_mom_def}  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$, {\bf we find}
\begin{align}
  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
%  & \leq 0.
\end{align}
{\bf
Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
negative angular momentum. If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We show 
below how this is possible through a forced temperature
gradient via the background torque. It is simplest to calculate
$T_\mathrm{BG}$ in the local approximation, which we review first.  
}

@@

\subsection{Local results}\label{local_approx}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. The
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc is  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
Note that in the {\bf strictly} local approximation, {\bf where all
  global effects are neglected}, only axisymmetric perturbations
($m=0$) can be unstable. % (having an imaginary $\sigma$).

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a {\bf non-axisymmetric} neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
waves.

{\bf At the \emph{co-rotation radius} $R_c$} the pattern speed matches
the fluid rotation,
\begin{align}
  \Omega(R_c) = \Omega_p.
\end{align}
\emph{Lindblad resonances} $R_L$ occurs where
\begin{align}
  \nu^2(R_L) = 1. 
\end{align}
Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
\begin{align}
  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
\end{align}
According to Eq. \ref{wavenumber}, purely wave-like solutions with
real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.

A detailed discussion of the properties of local density waves 
is given in \cite{shu91}. An important result, {\bf which holds for
  waves of all frequencies}, is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum.

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}
{\bf 
Here we show that the torque density acting on a 
local mode due to the background temperature gradient can be negative, which 
would enforce low-frequency modes, because they have negative angular momentum.  
}

The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
\end{align}
{\bf Inserting this into the expression for the background torque,
  Eq. \ref{baroclinic_torque}, we find}
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. \label{baroclinic_torque1}
\end{align}

@@

However, if the linear disturbance under consideration \emph{is} a
low-frequency mode, then it has negative angular
momentum. If it is trailing and $dc_s^2/dR<0$, as is typical
in astrophysical discs, then $T_\mathrm{BG}<0$ and
the background disc applies a negative torque on the disturbance,
which further decreases its angular momentum. This suggests the {\bf
  mode} 
amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
{\bf where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ for low-frequency modes in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}.} 
Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.}

{\bf Note that the derivation of Eq. \ref{baroclinic_torque1} and 
  Eq. \ref{theoretical_rate} do not 
  require the disc to be self-gravitating. Thus, destabilisation by the
  background torque is not directly associated with disc
  self-gravity. However, in order to evaluate 
  Eq. \ref{baroclinic_torque1} or Eq. \ref{theoretical_rate} in terms 
  of disc parameters (as done in \S\ref{fargo_m1}), we need to insert a 
  value of $k$, which may depend on disc    
  self-gravity (e.g. from Eq. \ref{wavenumber}).  
}

@@

{\bf
  We demonstrate the destabilising effect of a fixed temperature
  gradient using numerical simulations. The above discussion is
  generic for low-frequency non-axisymmetric modes, but for
  simulations we will consider specific examples.

  There are two parts to the destabilising mechanism: 
  the disc should support low-frequency modes, which is then
  destabilised by an imposed temperature gradient. The latter is
  straight forward to implement by adopting a locally isothermal
  equation of state as described in \S\ref{model}.  For the former, we
  consider discs with a  
  radially-structured Toomre $Q$ profile. We use local theory to show
  that such discs can trap 
  low-frequency one-armed ($m=1$) modes.  
  This is convenient because Eq. \ref{theoretical_rate} indicates that
  modes with small $m$ are
  more favourable for destabilisation. 
}

@@

\subsection{Codes}
{\bf We use three independent grid-based codes to simulate the above 
system.} We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.

>>>>

\section{Governing equations}\label{model} 
We consider an inviscid fluid disc of mass $M_d$ 
orbiting a central star of mass $M_*$. We will mainly examine 2D
(or razor-thin) discs in favour of numerical 
resolution, but have also carried out some 3D disc
simulations. Hence, for generality we describe the system in
3D, using both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the 
fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
$p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We impose a locally isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed $c_s$ is given by 
\begin{align}\label{sound-speed}
  % c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
\end{align}
where $c_{s0} = h R_0\Omega_k(R_0)$ and 
$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
$-q$ is the imposed temperature gradient since, for an ideal gas the
temperature is proportional to $c_s^2$. For convenience we will
  refer to $c_s^2$ as the disc temperature. The vertical disc scale-height is
defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
 disc with constant $H/R$.

@@

 We describe a key feature of locally isothermal discs that
  enables angular momentum exchange between small disturbances and the
  background disc through an imposed radial temperature gradient. This
  conclusion results from the consideration of angular momentum
  conservation within the framework of linear perturbation theory. 
  For simplicity, in this section we consider 2D discs.

@@

These linearised equations can be combined into a single
integro-differential equation eigenvalue problem. We defer a full
 numerical exploration of the linear problem to a future study. Here, we 
  discuss some
general properties of the linear perturbations.

@@

In Eq. \ref{angcons}, the background torque density $\tbg$ is 
\begin{align}\label{baroclinic_torque}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state in the perturbed disc.  We outline the
derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.

@@

The important consequence of the background torque is the possibility
of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
(negative) and $\tbg$ is also positive (negative),
then the local angular momentum density of the linear disturbance
 will further increase (decrease) with time. This implies the amplitude of the disturbance
may grow by exchanging angular momentum with the background
disc.

  We demonstrate instability for low-frequency modes 
  ($|\omega|\ll m\Omega$) by explicitly showing its 
  angular momentum density 
  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
  appropriate perturbations and radial temperature gradients.

\subsection{Angular momentum density of  non-axisymmetric low-frequency modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right].\label{ang_mom_def}  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$,  we find 
\begin{align}
  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R^*\xi_\phi\right)\right]\notag\\
  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
%  & \leq 0.
\end{align}

Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
negative angular momentum. If the mode loses (positive) angular momentum 
to the background, then we can expect instability. We show 
below how this is possible through a forced temperature
gradient via the background torque. It is simplest to calculate
$T_\mathrm{BG}$ in the local approximation, which we review first.

@@

\subsection{Local results}\label{local_approx}
In the local approximation, perturbations are assumed to vary rapidly
relative to any background gradients. The
dispersion relation for tightly-wound density 
waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
disc is  
\begin{align}\label{dispersion}
  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
Note that in the strictly local approximation, where all
  global effects are neglected, only axisymmetric perturbations
($m=0$) can be unstable. % (having an imaginary $\sigma$).

Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
\omega/m$ of a non-axisymmetric neutral mode 
, Eq. \ref{dispersion} can be solved 
for $|k|$, 
\begin{align}\label{wavenumber}
%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
 %     Q^2(1-\nu^2)}\right], 
|k| = k_c\left[1 \pm \sqrt{1 -
     Q^2(1-\nu^2)}\right], 
\end{align}
where 
 \begin{align}
   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
 \end{align}
 is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. In
Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
waves.

 At the \emph{co-rotation radius} $R_c$ the pattern speed matches
the fluid rotation,
\begin{align}
  \Omega(R_c) = \Omega_p.
\end{align}
\emph{Lindblad resonances} $R_L$ occurs where
\begin{align}
  \nu^2(R_L) = 1. 
\end{align}
Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
\begin{align}
  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
\end{align}
According to Eq. \ref{wavenumber}, purely wave-like solutions with
real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.

A detailed discussion of the properties of local density waves 
is given in \cite{shu91}. An important result, which holds for
  waves of all frequencies, is that
waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
waves outside co-rotation ($R>R_c$) have positive angular
momentum.

\subsection{Unstable interaction between low-frequency modes 
  and the background disc due to an imposed temperature gradient}

Here we show that the torque density acting on a 
local mode due to the background temperature gradient can be negative, which 
would enforce low-frequency modes, because they have negative angular momentum.

The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} can be neglected. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
\end{align}
 Inserting this into the expression for the background torque,
  Eq. \ref{baroclinic_torque}, we find 
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. \label{baroclinic_torque1}
\end{align}

@@

However, if the linear disturbance under consideration \emph{is} a
low-frequency mode, then it has negative angular
momentum. If it is trailing and $dc_s^2/dR<0$, as is typical
in astrophysical discs, then $T_\mathrm{BG}<0$ and
the background disc applies a negative torque on the disturbance,
which further decreases its angular momentum. This suggests the 
  mode  
amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadratic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
\begin{align}\label{theoretical_rate0}
  2\gamma \sim
  -\frac{dc_s^2}{dR}
  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
      \ii\xi_\phi|^2 \right]} 
  = -\frac{dc_s^2}{dR}
  \frac{k}{m\Omega},
%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
%   \xi_\phi|^2}. 
\end{align}
 where the second equality uses $\xi_\phi \simeq
2\ii \xi_R/m$ for low-frequency modes in the local approximation, as shown in  Appendix
\ref{horizontal_displacements}. 
Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
\end{align}
where we used $c_s\sim H\Omega\sim hR\Omega$. 
  Eq. \ref{theoretical_rate} suggests that perturbations with small
  radial length-scales ($kH\gtrsim m$) are most favourable for
  destabilisation. Taking the local approximation is then
  appropriate.

 Note that the derivation of Eq. \ref{baroclinic_torque1} and 
  Eq. \ref{theoretical_rate} do not 
  require the disc to be self-gravitating. Thus, destabilisation by the
  background torque is not directly associated with disc
  self-gravity. However, in order to evaluate 
  Eq. \ref{baroclinic_torque1} or Eq. \ref{theoretical_rate} in terms 
  of disc parameters (as done in \S\ref{fargo_m1}), we need to insert a 
  value of $k$, which may depend on disc    
  self-gravity (e.g. from Eq. \ref{wavenumber}).

@@

  We demonstrate the destabilising effect of a fixed temperature
  gradient using numerical simulations. The above discussion is
  generic for low-frequency non-axisymmetric modes, but for
  simulations we will consider specific examples.

  There are two parts to the destabilising mechanism: 
  the disc should support low-frequency modes, which is then
  destabilised by an imposed temperature gradient. The latter is
  straight forward to implement by adopting a locally isothermal
  equation of state as described in \S\ref{model}.  For the former, we
  consider discs with a  
  radially-structured Toomre $Q$ profile. We use local theory to show
  that such discs can trap 
  low-frequency one-armed ($m=1$) modes.  
  This is convenient because Eq. \ref{theoretical_rate} indicates that
  modes with small $m$ are
  more favourable for destabilisation.

@@

\subsection{Codes}
We use three independent grid-based codes to simulate the above 
system. We adopt computational units such that
$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
